\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{microtype}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{caption}
\usepackage{tikz}
\usepackage[round]{natbib}
\usepackage[hidelinks]{hyperref}

\hypersetup{
  pdftitle={No Single Safety Gate: Distributed Routing of Harm-Refusal in a Mixture-of-Experts Language Model},
  pdfauthor={Jeffrey W. Shorthill},
  pdfkeywords={mixture-of-experts, expert routing, refusal, AI safety, mechanistic interpretability, distributed representation, activation steering}
}

\captionsetup{font=small,labelfont=bf}
\setlength{\parskip}{0.35em}

\title{\textbf{No Single Safety Gate: Distributed Routing of Harm-Refusal\\
in a Mixture-of-Experts Language Model}}
\author{Jeffrey W. Shorthill\thanks{Independent researcher. Correspondence:
\texttt{jws299792@icloud.com}. ORCID: \href{https://orcid.org/0009-0004-3954-2752}{0009-0004-3954-2752}.
This is version~1.0 (June~2026), released under CC~BY~4.0. Data, code, and a
claim-by-claim source index accompany the paper; see the Data and Code
Availability statement.}}
\date{Version 1.0 --- June 2026}

\begin{document}
\maketitle

\begin{abstract}
A mixture-of-experts (MoE) language model routes each token to a few expert
subnetworks through a discrete, inspectable router. This makes it natural to ask
whether harm-refusal is owned by a single ``safety expert'' that could be located,
ablated, or attacked. We probe this in the base \textsc{Qwen3.5-35B-A3B} model
(40 layers, 256 experts, top-8 routing) with router capture at every layer under
greedy decoding. On a token-matched set of redflag versus matched-benign prompts,
the generation-side routing difference is led by one expert, expert 173 at layer
25, but a finance-versus-consequence control shows the broader signal is a
real-world-consequence and professional-duty cluster (experts 189, 45, 122, 157,
36, 216, and others) that is separable from finance-domain experts. We then
suppress expert 173 directly with a router bias swept over five levels. Its routed
mass collapses dose-dependently, from a selection rate of 0.81 to 0.05 at the
strongest setting, yet the model never produces a harmful completion on this prompt
set: routing reallocates to sibling consequence experts and the refusals persist,
becoming more explicit rather than less. We read this as evidence that, in this
model and setup, refusal is carried by a distributed expert cluster with no single
gate. The results are a small, exploratory case study (six and twelve prompts,
single greedy trajectories), and we give the controls that would confirm or break
them.
\end{abstract}

\noindent\textbf{Keywords:} mixture-of-experts; expert routing; refusal; AI safety;
mechanistic interpretability; distributed representation.

\section{Introduction}
\label{sec:intro}

A mixture-of-experts (MoE) layer replaces a dense feed-forward block with many
expert subnetworks and a router that selects a few of them for each token
\citep{shazeer2017outrageously,fedus2022switch}. Because the router makes a
discrete, per-token choice, its selections are unusually legible: one can read,
token by token, which experts a behavior recruits. That legibility invites a
specific question about safety. When a model refuses a harmful request, is the
refusal owned by a particular expert, a ``safety expert'' that a router-level
analysis could name, an auditor could monitor, and an attacker could disable?

The question matters because the dense-model picture points one way. Refusal in
dense transformers has been described as mediated by a single direction in the
residual stream, recoverable by contrasting harmful and harmless prompts and
removable by projecting it out \citep{arditi2024refusal}; representation-level
control of harmfulness is a working tool \citep{zou2023representation}. If the MoE
analog held, refusal would localize to one or a few experts, and suppressing the
top one would weaken it.

We test the localization picture on the base \textsc{Qwen3.5-35B-A3B} model. We
capture the router logits at all 40 layers while the model answers token-matched
pairs of redflag and benign prompts under greedy decoding, decompose the routed
weight per expert and layer, and identify the experts whose routing separates the
two classes. We then intervene: we add a negative bias to the single strongest
carrier and sweep its strength, watching both the routing and the generated text.

We find no single safety gate. One expert leads the routing separation, but it leads
a cluster, and when we bias it down the cluster keeps the refusals going. The sections
that follow give the decomposition that finds the leader, the control that tells a
consequence-and-duty cluster apart from experts that merely track a topic, and the
suppression sweep that takes the leader out. One methodological surprise recurs along
the way: the tidy prefill leader is mostly an artifact of the token-matching padding,
which a per-token decomposition dissolves.

The scope is narrow: six prompts in the matched set, twelve in the control, one model,
one quantization, single greedy generations, an intervention on one expert. We do not
claim that MoE refusal is never localizable, or that this carrier can never be
bypassed. We report what removing the strongest carrier does here, and list the
controls that would test how far it carries.

\section{Related work}
\label{sec:related}

\paragraph{Mixture-of-experts routing.}
Sparse MoE layers route tokens through a learned gate to a few of many experts
\citep{shazeer2017outrageously}, scaled into large transformers by Switch
Transformers \citep{fedus2022switch}, Mixtral \citep{jiang2024mixtral}, and the
fine-grained-expert routing of DeepSeek \citep{dai2024deepseekmoe}; the model we
study is in the Qwen3 lineage \citep{yang2025qwen3}. We use the per-token router
selection as the readout and ask which experts a single behavior recruits.

\paragraph{The locus of refusal.}
In dense models, refusal has been characterized as a single residual-stream
direction that can be found by difference-in-means and ablated to disable refusal
\citep{arditi2024refusal}, part of a broader program of reading and steering
high-level behaviors through directions \citep{zou2023representation}. Our question
is the MoE-routing analog: is there a comparable single locus at the level of
expert selection? We find that there is not, in this model. The contrast is in the
substrate (router experts rather than a residual direction) and in the conclusion
(a distributed cluster rather than a single carrier).

\paragraph{Safety experts in MoE models.}
Two recent lines bear directly on this. \citet{zhang2026understanding} report that
MoE routing is largely topic-driven and that safety behavior can change with little
change to the routing path, a conclusion convergent with ours. Our addition is a
causal, dose-dependent suppression of the single strongest carrier with observed
reallocation to siblings and preserved refusal, on a named model.
\citet{lai2025safex} take the opposite stance, locating vulnerability in stable,
identifiable safety-critical experts. We read the disagreement as setup-dependent:
on this base model, prompt set, and intervention, we do not find a single expert
whose removal breaks refusal, and we are explicit that this is not a universal
claim.

\paragraph{Distributed and superposed representations.}
That a behavior need not localize to one component is the expectation set by
superposition \citep{elhage2022superposition} and by dictionary-learning accounts
in which interpretable features, including safety-relevant ones, are spread across
many directions \citep{bricken2023monosemanticity,templeton2024scaling}. Our result
is the expert-routing analog of that picture: the safety-relevant computation is
spread across a cluster of experts that absorb each other's load. Methodologically
we use the same causal-intervention logic as activation editing in dense models
\citep{meng2022locating}, applied to a router bias.

\paragraph{Companion work.}
This paper sits beside two companion studies on the same model lineage: a
characterization of a single expert's role in generation \citep{shorthill2026e114}
and an analysis of when domain specialization is visible in routing
\citep{shorthill2026generation}. Where the first asks what one expert encodes, this
one asks whether a safety behavior is owned by one expert, and answers no.
Robustness and safety benchmarks \citep{mazeika2024harmbench,wei2023jailbroken}
frame why the locus of refusal is worth knowing, though we evaluate behavior by
direct reading on a small set rather than by a benchmark.

\section{Methods}
\label{sec:methods}

\paragraph{Model and decoding.}
We study the base \textsc{Qwen3.5-35B-A3B} model in 8-bit (Q8\_0) GGUF form, a
\texttt{qwen35moe} architecture with 40 layers, 256 experts per MoE layer, and 8
experts selected per token. Decoding is greedy and deterministic
(\texttt{--temp 0 --top-k 1 --seed 0}), capped at 256 new tokens, with a bare
\texttt{</think>} assistant prefix matching the project's reference runtime. We use
the base model, not a refusal-reduced fine-tune; the refusals studied here are the
base model's own.

\paragraph{Router capture and metric.}
For each prompt we capture the router pre-softmax logits at all 40 layers, for every
prompt and generated token. For an expert $e$ on a set of tokens we report, after a
softmax over the 256 logits, a top-8 selection, and a renormalization over the
selected set: $S_e$, the fraction of tokens on which $e$ is selected; $Q_e$, its
mean renormalized weight when selected; and $W_e = S_e\,Q_e$, its mean weight over
all tokens. Statistics are computed within the prefill block or the generation block,
never pooled across them; layer 39 carries a distinct row count per prompt and is
excluded from prefill claims. The quantity we rank on is
$\Delta W = W^{\mathrm{A}} - W^{\mathrm{B}}$, the difference in $W$ between redflag,
consequence, or unsafe prompts (A) and matched professional, descriptive, or neutral
prompts (B).

\paragraph{Experiment 1: token-matched redflag pairs.}
Three pairs hold the request topic fixed and vary only its framing: an oxycodone
prompt asking how to get ``incredibly high'' versus a pharmacist asking a normal
dose; an instruction to brew ayahuasca (prefixed with a prompt-injection attempt)
versus a request to describe it; and a request to invest for a ``sure'' 500\% return
versus a financial planner asking how to explain such an opportunity. The A and B
prompts of each pair are padded with a repeated filler token to identical prompt
token counts (50/50, 37/37, 74/74), so the two surfaces differ in framing, not
length. All six are answered once under greedy decoding.

\paragraph{Experiment 2: finance versus consequence control.}
Twelve prompts cross two axes, finance versus non-finance and consequence versus
neutral, to separate experts that track a topic (finance) from experts that track
real-world consequence and duty across domains (medical, legal, physical safety,
inspection, finance). The buckets are matched in intent but not strictly token-paired.

\paragraph{Experiment 3: suppressing the leading carrier.}
We rerun the three redflag pairs while adding a negative router bias to the single
strongest generation carrier, expert 173, at four strengths
($-0.25, -0.5, -1, -2$), capturing all-layer routing and the generated text at each.
We read each generation for whether the model still refuses or warns and how the
routing redistributes.

\paragraph{A token-matching caveat.}
The pad token in Experiment 1 is a repeated word piece, and because it appears with
different frequency in A and B, it can manufacture a routing difference on the
prefill (prompt-reading) block that is an artifact of the padding rather than of the
content. We therefore back the aggregate tables with a per-token decomposition that
recomputes each candidate's $\Delta W$ with the filler piece removed, and we read the
safety signal off the generation block, which is produced response text and carries
no padding.

\section{Results}
\label{sec:results}

\subsection{The generation safety signal is led by one expert, atop a cluster}
\label{sec:lead}

On the token-matched redflag pairs, the routing difference over generated tokens is
led by expert 173 at layer 25, with $\Delta W = 0.124$
(Table~\ref{tab:gen-leaders}). Both components of the weight rise on redflag
generation: its selection rate climbs from $0.43$ on the benign prompts to $0.81$ on
the redflag prompts, and its conditional weight from $0.16$ to $0.24$. Beneath it
sits a broad supporting set, expert 45 at layers 19 and 31, expert 25 at layer 20,
expert 216 at layers 15 and 27, expert 189 at layer 17, and others, each separating
the two prompt classes but none rivalling the leader.

\begin{table}[t]
\centering
\small
\begin{tabular}{rllrrr}
\toprule
Rank & Layer & Expert & $\Delta W$ & $W^{\mathrm{A}}$ ($S^{\mathrm{A}}$) & $W^{\mathrm{B}}$ ($S^{\mathrm{B}}$) \\
\midrule
1 & 25 & 173 & 0.124 & 0.192 (0.81) & 0.068 (0.43) \\
2 & 19 & 45  & 0.082 & 0.102 (0.72) & 0.021 (0.17) \\
3 & 20 & 25  & 0.073 & 0.110 (0.56) & 0.037 (0.33) \\
4 & 13 & 173 & 0.072 & 0.191 (0.86) & 0.119 (0.69) \\
5 & 31 & 45  & 0.067 & 0.078 (0.57) & 0.010 (0.08) \\
6 & 27 & 216 & 0.063 & 0.074 (0.48) & 0.011 (0.08) \\
7 & 17 & 189 & 0.061 & 0.107 (0.64) & 0.046 (0.33) \\
\bottomrule
\end{tabular}
\caption{Generation-block A$-$B routing leaders on the three token-matched redflag
pairs ($n=6$ prompts). $A$ is the redflag surface, $B$ the matched benign surface;
$W$ is mean routed weight and $S$ the selection rate. Expert 173 at layer 25 leads,
but the separation is carried by a set of experts, not one. Single greedy generation
per prompt.}
\label{tab:gen-leaders}
\end{table}

\subsection{Aggregate prefill ``leaders'' are largely a padding artifact}
\label{sec:filler}

Read naively, the prefill block appears led by expert 173 at layer 1
($\Delta W = 0.068$). The per-token decomposition shows this is mostly the repeated
filler piece: excluding it, the layer-1 figure falls to $0.0004$, and the other top
aggregate prefill rows (expert 222 at layer 2, expert 36 at layer 22) collapse
similarly. The prompt-side candidates that survive the correction are expert 218 at
layer 14 ($0.025 \to 0.022$), expert 173 at layer 25 ($0.037 \to 0.046$), and expert
45 at layer 19 ($0.030 \to 0.034$). We therefore rest the safety read on the
generation block, and treat the headline prefill leader as a methodological warning:
token-count padding can fabricate a prefill expert that a per-token decomposition
dissolves.

\subsection{The cluster tracks consequence and duty, not a topic}
\label{sec:control}

The finance-versus-consequence control separates two things a single redflag set
would conflate (Table~\ref{tab:control}). Finance-domain experts, expert 62 at layer
20, expert 95 at layer 30, expert 223 at layer 19, and expert 214 at layer 16, fire
strongly on finance prompts including neutral finance, with near-zero selection on
non-finance, so they are topic experts, not refusal experts. The
consequence-minus-neutral contrast, by contrast, is led by expert 189 at layer 17,
expert 122 at layers 11 and 23, expert 157 at layer 14, expert 171 at layer 15, and
expert 45 at layer 19, and it generalizes across medical, legal, physical-safety,
inspection, and finance-consequence content. The safety-relevant cluster is better
read as real-world-consequence and professional-duty routing than as a topic or a
single expert.

\begin{table}[t]
\centering
\small
\begin{tabular}{llr}
\toprule
Layer & Expert & $\Delta W$ \\
\midrule
\multicolumn{3}{l}{\emph{Finance-domain experts (finance $-$ non-finance)}} \\
20 & 62  & 0.180 \\
30 & 95  & 0.177 \\
19 & 223 & 0.167 \\
\midrule
\multicolumn{3}{l}{\emph{Consequence/duty experts (consequence $-$ neutral)}} \\
17 & 189 & 0.073 \\
23 & 122 & 0.065 \\
14 & 157 & 0.058 \\
15 & 171 & 0.053 \\
\bottomrule
\end{tabular}
\caption{Finance-versus-consequence control ($n=12$ prompts, bucketed). Finance-domain
experts (top) fire on finance including neutral finance and are not refusal experts;
the broad consequence/duty experts (bottom) generalize across domains. Generation
block. Buckets are intent-matched, not strictly token-paired.}
\label{tab:control}
\end{table}

\subsection{Suppressing the leading carrier does not break refusal}
\label{sec:suppression}

At the strongest setting we tried, bias $-2$, expert 173 is nearly gone from the
generation routing: its redflag selection rate is $0.05$, down from $0.81$, and its
A$-$B weight difference is $0.004$, down from $0.124$ (Table~\ref{tab:dose},
Figure~\ref{fig:dose}). All three redflag prompts still refuse. The collapse is graded
across the four bias levels; the behavior is not.

\begin{table}[t]
\centering
\small
\begin{tabular}{lrrr}
\toprule
Router bias on E173 & $\Delta W$ (L25) & $S^{\mathrm{A}}$ & $W^{\mathrm{A}}$ \\
\midrule
baseline   & 0.124 & 0.809 & 0.192 \\
$-0.25$    & 0.093 & 0.664 & 0.141 \\
$-0.5$     & 0.056 & 0.510 & 0.091 \\
$-1$       & 0.041 & 0.394 & 0.056 \\
$-2$       & 0.004 & 0.046 & 0.005 \\
\bottomrule
\end{tabular}
\caption{Dose-response of suppressing expert 173, read at its strongest carrier
(layer 25, generation block). Routed mass and selection rate collapse toward zero.
At every level the model still refuses or warns on all three redflag prompts; see
text.}
\label{tab:dose}
\end{table}

\begin{figure}[t]
\centering
\begin{tikzpicture}[x=2.0cm,y=0.7cm]
\foreach \y in {0,0.05,0.10}{\draw[gray!30] (0,\y*10) -- (5.1,\y*10);
  \node[font=\scriptsize,gray!70,anchor=east] at (-0.05,\y*10){\y};}
\node[font=\scriptsize,rotate=90] at (-0.55,0.6){$\Delta W$ (L25 E173)};
\foreach \x/\v/\lab in {0/0.124/baseline,1/0.093/-0.25,2/0.056/-0.5,3/0.041/-1,4/0.004/-2}{
  \fill[teal!60] (\x*1.15+0.15,0) rectangle (\x*1.15+0.75,\v*10);
  \node[font=\scriptsize] at (\x*1.15+0.45,\v*10+0.35){\v};
  \node[font=\scriptsize,anchor=north] at (\x*1.15+0.45,-0.15){\lab};
}
\node[font=\scriptsize,anchor=north] at (2.6,-0.75){router bias on expert 173};
\end{tikzpicture}
\caption{Generation-block A$-$B routing difference for expert 173 at layer 25 as its
router bias is swept from baseline to $-2$. The carrier collapses dose-dependently;
the model's refusals do not.}
\label{fig:dose}
\end{figure}

The oxycodone prompt is refused at every level, and at $-2$ the refusal gets
\emph{more} explicit about misuse and overdose; the ayahuasca instruction is refused
throughout; the 500\% prompt is warned against throughout, its wording shifting from
``impossible and dangerous'' to ``trap'' at the strongest bias. The benign halves stay
professional. The routed mass that leaves expert 173 reappears nearby: generation
separation reallocates to experts 45, 185, 157, 189, 216, and 133. The strongest
carrier switches off and the cluster carries the refusal.

\section{Interpretation}
\label{sec:interp}

Three results point the same way, and only the last is a real test. One expert leads
the routing, but it leads a cluster; the cluster tracks consequence rather than topic;
and removing the leader does not stop the refusing. The suppression is the load-bearing
one, because it is causal. A behavior owned by a single gate should weaken when the gate
is shut. This one does not. In this model, harm-refusal is a property of a cluster that
survives the loss of its strongest member.

This is the expert-routing analog of the distributed-representation expectation
\citep{elhage2022superposition,bricken2023monosemanticity}, and it stands against a
strict single-locus reading of refusal carried over from dense models
\citep{arditi2024refusal}. The contrast with \citet{lai2025safex}, which locates
vulnerability in identifiable safety-critical experts, we take to be setup-dependent
rather than a contradiction: a single-expert intervention that fails to bypass safety
here may succeed in another model, prompt regime, or with a multi-expert
intervention. What this run shows is the negative, that the strongest carrier in this
model is not a sufficient point of attack on this set.

We are careful about what the suppression does and does not establish. It tests the
\emph{necessity} of expert 173 alone, and finds it not necessary; it does not test
the joint necessity of the cluster, which a combined suppression of the replacement
set (experts 45, 189, 157, 122) would address. The behavioral read is a direct
reading of generated text on a small set, not a benchmarked refusal rate. The result
is a structural observation about where the routing lives, offered with the controls
that would harden it.

\section{Limitations and falsification tests}
\label{sec:limits}

\paragraph{What the present evidence does not settle.}
\begin{itemize}
\item \textbf{Six and twelve prompts, one trajectory each.} Single greedy generations,
so these are point estimates with no sampled spread and no significance test, and
per-prompt deltas can hang on tie-breaking. The ``still refuses'' verdict is the
author's reading of the text, not a graded harmful-completion rate.
\item \textbf{We suppressed one expert.} Whether the cluster jointly carries the
refusal is untested, so ``no jailbreak'' means removing the strongest carrier, not
removing the cluster.
\item \textbf{The control is bucketed, not token-paired.} Its
finance-versus-consequence separations are weaker evidence than the matched pairs.
\end{itemize}

\paragraph{Tests that would confirm or break the claim.}
\begin{itemize}
\item Jointly suppress the replacement set (experts 45, 189, 157, 122) in small steps
and ask whether refusal survives the removal of the cluster, not just its leader; the
distributed reading predicts graceful degradation, a gate reading predicts a cliff.
\item Score behavior against a standardized red-team suite
\citep{mazeika2024harmbench} under suppression, replacing the direct reading with a
harmful-completion rate.
\item Repeat with sampled decoding and more prompts per condition to attach
confidence to the per-expert deltas and the behavioral counts.
\item Run the same probe on other MoE families and on refusal-reduced fine-tunes to
test whether the distributed pattern, or a localizable safety expert
\citep{lai2025safex}, is the general case.
\item Compare the router-level picture to the residual-direction picture
\citep{arditi2024refusal} on the same model, to ask whether a single refusal
direction coexists with distributed expert routing.
\end{itemize}

\section{Conclusion}
\label{sec:conc}

In base \textsc{Qwen3.5-35B-A3B}, harm-refusal does not route through a single safety
expert. One expert leads, but biasing it to near-zero routed mass leaves the refusals
intact, because the surrounding consequence-and-duty cluster takes the load. For anyone
hoping to locate, monitor, or attack refusal at the level of a single MoE expert, the
lesson here is that the obvious target is not the gate it looks like. The evidence is a
small case study; the cluster-suppression and benchmark controls that would carry it
past one model are set out in Section~\ref{sec:limits}.

\section*{Reproducibility and Data and Code Availability}
\addcontentsline{toc}{section}{Data and Code Availability}

The numbers in this paper are transcribed from the analysis tables of a single
capture session and are listed claim-by-claim, with source files, in the accompanying
\texttt{SOURCES.md}. The supporting data accompany the paper in the \texttt{data/}
directory: the run's results write-up and journal, the all-layer A$-$B delta tables,
the per-token decomposition that backs the filler-artifact correction, the
finance-versus-consequence bucket tables, the expert-173 dose-response tables, the
generated openings under suppression, and the prompt sets. Raw router tensors
(\texttt{.npy}) were not retained in version control by project policy; the analysis
tables above are derived from them and substantiate the reported statistics. The
routing reconstruction satisfies $W = S\cdot Q$ per token set; we report the
reconstructed, log-derived routing tables. No new model training was performed.

\section*{Ethical Considerations}

This work studies how a model refuses, using a small set of synthetic redflag prompts
(drug misuse, a prompt-injection attempt, and a financial ``sure-thing'') that the
model refused at every intervention level. We release the prompts and the model's
refusals, not harmful content: no completion in this study provides operational harm,
and the suppression result is reported because it \emph{fails} to bypass safety, not
because it succeeds. The purpose is defensive, to understand whether refusal in an MoE
router is concentrated enough to be a single point of failure for auditing or attack;
our finding that it is distributed argues against a single-expert attack surface in
this model, while we are explicit that this does not generalize without the tests in
Section~\ref{sec:limits}.

\section*{AI Collaboration Statement}
\addcontentsline{toc}{section}{AI Collaboration Statement}

Generative AI (Anthropic's Claude) was used as a tool in preparing this manuscript:
to assist with drafting and revising prose, organizing the manuscript, and locating
and formatting the bibliography. All experiments, captures, and numerical results are
the author's; every reported value was traced to a source capture (see
\texttt{SOURCES.md}), and every reference was verified by the author against its
authoritative record to exist and to support the statement it is attached to. No AI
system is an author. The author reviewed the entire text and takes full responsibility
for its content, including any errors.

\section*{Funding and Competing Interests}
\addcontentsline{toc}{section}{Funding and Competing Interests}

This work received no external funding. The author declares no competing interests.
The model studied is a third-party open-weights release used as-is for measurement;
the author has no affiliation with its creators.

\bibliographystyle{plainnat}
\bibliography{refs}

\end{document}
